---
title: "Workshop 2: Spatial Econometrics & Road Networks"
subtitle: "Does location matter? Spatial patterns, road access, and urban regression"
author: "Jem Marie Nario"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-copy: true
    highlight-style: github
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    embed-resources: true
    smooth-scroll: true
    css: styles.css
execute:
  echo: true
  warning: false
  message: false
  cache: false
editor:
  markdown:
    wrap: 72
---

## Overview

In Workshop 1, we mapped rent values and human capital across Metro
Manila. We treated each city as an **independent** observation — as if
Makati and Mandaluyong had nothing to do with each other.

But cities are not islands. What happens in one city often spills over
into its neighbours. **Spatial econometrics** gives us tools to detect
and model these spillovers.

By the end of this workshop, you will be able to:

1.  Explain what spatial autocorrelation is and why it matters
2.  Compute and interpret **Moran's I** — the standard test for spatial
    autocorrelation
3.  Download and visualise the **road network** of Metro Manila using
    `osmdata` and `dodgr`
4.  Measure how accessible each city is to the Makati CBD via roads
5.  Run a **spatial lag model** and compare it to a standard OLS
    regression

::: callout-note
**Packages used in this workshop**\
`sf`, `spdep`, `spatialreg`, `osmdata`, `dodgr`, `ggplot2`, `dplyr`\
All can be installed with `install.packages()` — see Section 1.
:::

------------------------------------------------------------------------

## 1 · Setup

### 1.1 Install and load packages

```{r}
#| label: setup

# install.packages(c("spdep", "spatialreg", "osmextract", "dodgr",
#                    "sf", "ggplot2", "dplyr", "ggrepel",
#                    "scales", "viridis", "patchwork", "geodata"))

library(tidyverse)
library(sf)
library(geodata)
library(spdep)       # spatial weights and Moran's I
library(spatialreg)  # spatial lag and error models
library(osmextract)  # download OpenStreetMap data via Geofabrik
library(dodgr)       # road network distances
library(ggplot2)
library(ggrepel)
library(scales)
library(viridis)
library(patchwork)
```

### 1.2 Load the data from Workshop 1

We reuse the same NCR city dataset and shapefile from Workshop 1.

```{r}
#| label: load-data

load("data/ncr_city_training.RData")   # loads 'ncr_city'

# Rebuild the shapefile and join — same steps as Workshop 1
ph_adm2    <- gadm(country = "PHL", level = 2, path = "data/")
ph_adm2_sf <- st_as_sf(ph_adm2)

ncr_sf <- ph_adm2_sf %>%
  filter(NAME_1 == "Metropolitan Manila") %>%
  mutate(city = case_when(
    NAME_2 == "Las Piñas"     ~ "Las Pinas",
    NAME_2 == "Parañaque"     ~ "Paranaque",
    NAME_2 == "Kalookan City" ~ "Caloocan",
    NAME_2 == "Makati City"   ~ "Makati",
    NAME_2 == "Pasay City"    ~ "Pasay",
    NAME_2 == "Pasig City"    ~ "Pasig",
    TRUE                      ~ NAME_2
  ))

ncr_map <- ncr_sf %>%
  left_join(ncr_city, by = "city")

ncr_centroids <- ncr_map %>%
  st_centroid() %>%
  mutate(
    lon = st_coordinates(.)[, 1],
    lat = st_coordinates(.)[, 2]
  )

cat("Cities loaded:", nrow(ncr_map), "\n")
```

------------------------------------------------------------------------

## 2 · What Is Spatial Autocorrelation?

### 2.1 The core idea

When we run a regression in standard economics, we assume our
observations are **independent** of each other. This is often reasonable
for households — but it is usually *wrong* for cities.

Cities that are next to each other tend to be **similar**:

- A high-rent city is usually surrounded by other high-rent cities
- Workers commute across city borders, so wages spill over
- Infrastructure (roads, trains) links nearby cities together

This tendency for nearby places to look alike is called **spatial
autocorrelation**. Ignoring it can bias our regression results.

::: callout-note
**Tobler's First Law of Geography**\
*"Everything is related to everything else, but near things are more
related than distant things."*\
— Waldo Tobler (1970)

This is the founding principle of spatial econometrics.
:::

### 2.2 Positive vs negative autocorrelation

| Pattern | What it looks like | Metro Manila example |
|------------------------|------------------------|------------------------|
| **Positive** | High values cluster with high; low with low | High-rent cities (Makati, BGC area) cluster together |
| **Negative** | High values border low values (checkerboard) | Rare in urban rent data |
| **None (random)** | No spatial pattern | What standard OLS assumes |

------------------------------------------------------------------------

## 3 · Building a Spatial Weights Matrix

To measure spatial relationships, we need to define which cities are
**neighbours**. This is captured in a **spatial weights matrix** W.

### 3.1 Queen contiguity

The simplest rule: two cities are neighbours if they **share a border or
a corner** (like a queen in chess).

```{r}
#| label: spatial-weights

# Build queen contiguity neighbours
ncr_nb <- poly2nb(ncr_map, queen = TRUE)

# Convert to a weights list (row-standardised)
# row-standardised means each row sums to 1
ncr_w <- nb2listw(ncr_nb, style = "W")

# Summary
summary(ncr_nb)
```

```{r}
#| label: plot-neighbours
#| fig-cap: "Queen contiguity neighbours — NCR cities"

# Extract centroid coordinates for plotting the neighbour links
coords <- st_coordinates(st_centroid(ncr_map))

plot(st_geometry(ncr_map), border = "grey60", col = "grey95",
     main = "Spatial Neighbours (Queen Contiguity)")
plot(ncr_nb, coords, add = TRUE, col = "#c0392b", lwd = 1.2)
text(coords, labels = ncr_map$city, cex = 0.55, col = "grey20")
```

::: callout-tip
**Reading the neighbour plot**\
Each red line connects two cities that are treated as neighbours in our
spatial model. Makati, which is centrally located and small, will have
many neighbours. Peripheral cities like Las Pinas or Valenzuela will
have fewer.
:::

------------------------------------------------------------------------

## 4 · Moran's I — Testing for Spatial Autocorrelation

**Moran's I** is the standard statistic for detecting spatial
autocorrelation. Think of it as the spatial version of a correlation
coefficient.

| Moran's I value | Interpretation                               |
|-----------------|----------------------------------------------|
| Close to **+1** | Strong positive autocorrelation (clustering) |
| Close to **0**  | No spatial pattern (random)                  |
| Close to **−1** | Negative autocorrelation (checkerboard)      |

### 4.1 Moran's I for rent

```{r}
#| label: morans-i-rent

moran_rent <- moran.test(ncr_map$mean_rentval, ncr_w)
moran_rent
```

```{r}
#| label: moran-plot-rent
#| fig-cap: "Moran scatterplot — mean rent value"

# The Moran scatterplot shows spatially lagged rent (y) vs own rent (x)
# A positive slope = positive spatial autocorrelation
moran.plot(
  as.vector(scale(ncr_map$mean_rentval)),
  ncr_w,
  labels  = ncr_map$city,
  xlab    = "Rent (standardised)",
  ylab    = "Spatially lagged rent (neighbours' average)",
  main    = "Moran Scatterplot — Rent Value"
)
```

```{r}
# Identify the LH quadrant cities (potential gentrification pressure)
z <- as.vector(scale(ncr_map$mean_rentval))
lag_z <- lag.listw(ncr_w, z)

ncr_map %>%
  st_drop_geometry() %>%
  mutate(z = z, lag_z = lag_z) %>%
  filter(z < 0, lag_z > 0) %>%
  select(city, mean_rentval, z, lag_z) %>%
  arrange(desc(lag_z))
```

::: callout-important
**Interpreting the Moran scatterplot**\
The x-axis is each city's own standardised rent. The y-axis is the
*average rent of its neighbours*. A positive slope confirms that
high-rent cities tend to be surrounded by other high-rent cities.

The four quadrants tell you:

- **Top-right (HH):** high rent, high-rent neighbours → hot spots
- **Bottom-left (LL):** low rent, low-rent neighbours → cold spots
- **Top-left (LH):** low rent but high-rent neighbours → potential
  gentrification pressure
- **Bottom-right (HL):** high rent but low-rent neighbours → isolated
  islands
:::

In our NCR data, **Makati and San Juan** fall in the HL quadrant —
they have the highest own rent but their neighbours' average is near
zero, making them isolated high-rent islands rather than centres of a
broader high-rent cluster. The LH quadrant is more policy-relevant:
**Pateros, Pasay, Manila, Taguig, and Quezon City** all have
below-average rent yet sit next to relatively expensive neighbours.
These cities face the classic gentrification pressure — rising land
values spilling in from adjacent areas without a corresponding rise in
local incomes.


### 4.2 Moran's I for college graduation

```{r}
#| label: morans-i-college

moran_college <- moran.test(ncr_map$pct_college_grad, ncr_w)
moran_college
```

```{r}
#| label: moran-plot-college
#| fig-cap: "Moran scatterplot — college graduation rate"

moran.plot(
  as.vector(scale(ncr_map$pct_college_grad)),
  ncr_w,
  labels = ncr_map$city,
  xlab   = "College grads % (standardised)",
  ylab   = "Spatially lagged college grads (neighbours' average)",
  main   = "Moran Scatterplot — College Graduation Rate"
)
```
```{r}
#| label: college-quadrants

z_col   <- as.vector(scale(ncr_map$pct_college_grad))
lag_col <- lag.listw(ncr_w, z_col)

ncr_map %>%
  st_drop_geometry() %>%
  mutate(z = z_col, lag_z = lag_col) %>%
  filter(z < 0, lag_z > 0) %>%
  select(city, pct_college_grad, z, lag_z) %>%
  arrange(desc(lag_z))
```

::: callout-important
**Interpreting the Moran scatterplot**\
The x-axis is each city's own standardised college graduation rate.
The y-axis is the *average graduation rate of its neighbours*. A
positive slope confirms that high-education cities cluster together
spatially.

The four quadrants tell you:

- **Top-right (HH):** high education, high-education neighbours → knowledge clusters
- **Bottom-left (LL):** low education, low-education neighbours → persistent skills gaps
- **Top-left (LH):** low education but high-education neighbours → spillover potential
- **Bottom-right (HL):** high education but low-education neighbours → isolated skills hubs
:::

Only **Mandaluyong** falls in the LH quadrant — it has a slightly
below-average college graduation rate yet sits next to high-education
neighbours (Makati, San Juan, Pasig). This makes it the most likely
candidate for human capital spillovers: workers and firms in
Mandaluyong benefit from proximity to a highly skilled labour pool
without the city itself having produced those graduates locally.


------------------------------------------------------------------------

## 5 · Road Network Data with `osmdata`

So far we have used straight-line distance from Makati CBD. But people
do not travel in straight lines — they use roads. In this section, we
download the actual Metro Manila road network from **OpenStreetMap
(OSM)**.

### 5.1 What is OpenStreetMap?

OpenStreetMap (OSM) is a free, community-built map of the world —
similar to Google Maps but open and downloadable. The `osmextract`
package downloads pre-built OSM extracts from Geofabrik — a mirror
that packages OSM data by country — directly into R as an `sf` object.
This is faster and more reliable than querying the live Overpass API.

### 5.2 Download the road network

::: callout-warning
**First run downloads ~30–80MB from Geofabrik.**
This takes 1–2 minutes once, then saves to `data/roads_ncr.gpkg`.
Every subsequent render reads from disk in under 5 seconds —
no internet connection required after that.
:::


```{r}
#| label: download-osm-roads
#| cache: true

if (!file.exists("data/roads_ncr.gpkg")) {
  ph_roads <- oe_get(
    "Philippines",
    layer              = "lines",
    query              = "SELECT * FROM lines WHERE highway IN
                         ('motorway','trunk','primary','secondary','tertiary')",
    download_directory = "data/"
  )

  bbox_ncr <- st_bbox(
    c(xmin = 120.92, ymin = 14.38, xmax = 121.10, ymax = 14.76),
    crs = st_crs(4326)
  ) %>% st_as_sfc()

  roads <- ph_roads[st_intersects(ph_roads, bbox_ncr, sparse = FALSE), ]
  st_write(roads, "data/roads_ncr.gpkg", delete_dsn = TRUE)
} else {
  roads <- st_read("data/roads_ncr.gpkg")
}

cat("Road segments loaded:", nrow(roads), "\n")
```

### 5.3 Visualise the road network

```{r}
#| label: plot-roads
#| fig-cap: "Metro Manila road network by road type (OSM)"

# Classify road type for styling
roads <- roads %>%
  mutate(road_class = case_when(
    highway %in% c("motorway", "trunk") ~ "Expressway / Trunk",
    highway == "primary"                ~ "Primary",
    highway == "secondary"              ~ "Secondary",
    highway == "tertiary"               ~ "Tertiary",
    TRUE                                ~ "Other"
  ))

ggplot() +
  geom_sf(data = ncr_map, fill = "#f5f5f0", color = "white",
          linewidth = 0.6) +
  geom_sf(data = filter(roads, road_class == "Other"),
          color = "#cccccc", linewidth = 0.15, alpha = 0.6) +
  geom_sf(data = filter(roads, road_class == "Tertiary"),
          color = "#aaaaaa", linewidth = 0.25) +
  geom_sf(data = filter(roads, road_class == "Secondary"),
          color = "#888888", linewidth = 0.4) +
  geom_sf(data = filter(roads, road_class == "Primary"),
          color = "#c0392b", linewidth = 0.65) +
  geom_sf(data = filter(roads, road_class == "Expressway / Trunk"),
          color = "#e67e22", linewidth = 0.9) +
  labs(
    title   = "Metro Manila Road Network",
    subtitle = "Orange = expressway/trunk · Red = primary · Grey = secondary/tertiary",
    caption = "Source: © OpenStreetMap contributors"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5,
                                 margin = margin(b = 4)),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
    plot.caption  = element_text(color = "grey60", size = 7, hjust = 0.5)
  )
```

------------------------------------------------------------------------

## 6 · Road Network Accessibility with `dodgr`

Straight-line distance from Makati CBD (our `dist_cbd_km` variable)
ignores road layout. A city 5 km away by air may be 12 km by road if
there is a river or a wall in between.

`dodgr` (**D**istances **O**n **D**irected **G**raphs in **R**) computes
actual travel distances along a road network.

### 6.1 Build the road graph

```{r}
#| label: build-road-graph

# dodgr needs OSM data as a data frame — weight_streetnet() converts it
road_graph <- weight_streetnet(
  roads,
  wt_profile = "motorcar"   # weight by car travel time
)

cat("Road graph edges:", nrow(road_graph), "\n")
```

### 6.2 Define origin and destination points

```{r}
#| label: od-points

# Origin: each city centroid
origins <- ncr_centroids %>%
  st_transform(4326) %>%             # ensure WGS84
  select(city, lon, lat) %>%
  st_drop_geometry()

# Destination: Makati CBD (Ayala Avenue area)
makati_cbd <- data.frame(
  lon = 121.0244,
  lat = 14.5547
)
```

### 6.3 Compute road distances

```{r}
#| label: road-distances
#| cache: true

# dodgr_dists() returns a matrix of distances in metres
# from each origin to each destination
dist_matrix <- dodgr_dists(
  graph = road_graph,
  from  = as.matrix(origins[, c("lon", "lat")]),
  to    = as.matrix(makati_cbd)
)

# Attach road distance (km) back to city data
origins$road_dist_km <- as.numeric(dist_matrix) / 1000

ncr_city <- ncr_city %>%
  left_join(select(origins, city, road_dist_km), by = "city")

cat("Road distances computed. Preview:\n")
origins %>%
  arrange(road_dist_km) %>%
  select(city, road_dist_km) %>%
  print()
```

### 6.4 Compare road vs straight-line distance

```{r}
#| label: road-vs-straight
#| fig-cap: "Road distance vs straight-line distance to Makati CBD"

# Remove road_dist_km if it already exists to avoid duplicate columns
ncr_city <- ncr_city %>% select(-any_of("road_dist_km"))

ncr_city <- ncr_city %>%
  left_join(select(origins, city, road_dist_km), by = "city")

ggplot(ncr_city, aes(x = dist_cbd_km, y = road_dist_km)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey60") +
  geom_point(color = "#2c3e50", size = 3, alpha = 0.85) +
  geom_text_repel(aes(label = city), size = 3.2, color = "grey30",
                  max.overlaps = 20) +
  labs(
    title    = "Road Distance vs Straight-Line Distance",
    subtitle = "Points above the dashed line take longer by road than by air",
    x        = "Straight-line distance (km)",
    y        = "Road network distance (km)",
    caption  = "Sources: GADM, OSM via dodgr"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

::: callout-tip
**Reading this chart**\
The dashed line is the 45° line — it shows where road distance equals
straight-line distance. Every city falls *above* this line (roads are
always longer than the crow flies). Cities furthest above the line have
the worst road connectivity relative to their geographic position.
:::

The most striking outlier is **Caloocan** — it sits about 11 km from
Makati by air but requires over 23 km by road, the largest gap in NCR.
This reflects the barrier effect of Quezon City's layout and the lack
of direct north-south expressways through the inner city. **Valenzuela
and Muntinlupa** also show large gaps, both being peripheral cities
where the road network routes traffic around rather than through the
urban core. At the other end, **Mandaluyong and San Juan** sit close to
the 45° line — their road distance is nearly equal to straight-line
distance, meaning they are exceptionally well-connected relative to
their position. This plot is a simple but powerful illustration of why
straight-line distance is a poor proxy for market access in a city
shaped by rivers, rail lines, and unplanned road layouts.


### 6.5 Map road accessibility

```{r}
#| label: map-accessibility
#| fig-cap: "Road network distance to Makati CBD (km) — NCR cities"

ncr_map <- ncr_map %>%
  left_join(select(ncr_city, city, road_dist_km), by = "city")

ncr_centroids <- ncr_centroids %>%
  left_join(select(ncr_city, city, road_dist_km), by = "city")

ggplot(ncr_map) +
  geom_sf(aes(fill = road_dist_km), color = "white", linewidth = 0.8) +
  geom_text_repel(
    data          = ncr_centroids,
    aes(x = lon, y = lat, label = city),
    size          = 3,
    color         = "grey20",
    max.overlaps  = 20,
    segment.size  = 0.3,
    segment.color = "grey60"
  ) +
  scale_fill_distiller(
    palette   = "RdYlGn",
    direction = -1,
    name      = "Road dist\nto CBD (km)"
  ) +
  labs(
    title   = "Road Accessibility to Makati CBD",
    subtitle = "Darker green = closer by road · Darker red = further by road",
    caption  = "Source: OSM via dodgr + GADM"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5,
                                 margin = margin(b = 4)),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
    plot.caption  = element_text(color = "grey60", size = 7, hjust = 0.5),
    legend.position = "right"
  )
```

------------------------------------------------------------------------

## 7 · Spatial Regression Models

We now have three things:

1.  **An outcome** — mean rent value (`mean_rentval`)
2.  **An explanatory variable** — road distance to CBD (`road_dist_km`)
3.  **A spatial weights matrix** — who is neighbours with whom (`ncr_w`)

Let's put these together and see whether spatial spillovers matter.

### 7.1 Step 1 — Baseline OLS

Start with a standard regression, ignoring spatial structure:

```{r}
#| label: ols-model

ols <- lm(log(mean_rentval) ~ road_dist_km + pct_college_grad,
          data = ncr_map)
summary(ols)
```

::: callout-note
**Why log(rent)?**\
Taking the log of rent makes the model multiplicative: a 1 km increase
in road distance changes rent by a *percentage* rather than a fixed peso
amount. This is the standard approach in urban economics (the
Alonso–Muth–Mills gradient).
:::

### 7.2 Step 2 — Test residuals for spatial autocorrelation

After running OLS, we test whether the *errors* are spatially
correlated. If they are, OLS gives us biased standard errors.

```{r}
#| label: residual-moran

# Moran's I on OLS residuals
moran_resid <- lm.morantest(ols, ncr_w)
moran_resid
```

::: callout-important
**Decision rule**\
If the p-value from `lm.morantest()` is below 0.05, the OLS residuals
are spatially autocorrelated → OLS is not appropriate → use a spatial
model.
:::

The Moran's I on the OLS residuals is 0.104 with a p-value of 0.065.
This is a borderline result — it does not clear the conventional 0.05
threshold, so we cannot firmly reject the null of no spatial
autocorrelation in the residuals. However, with only 17 observations
(NCR cities), the test has very low statistical power, meaning genuine
spatial structure can easily go undetected. The positive Moran's I
value still suggests a tendency for residuals to cluster spatially, so
it is worth proceeding with the spatial models as a robustness check
rather than treating OLS as definitively adequate.


### 7.3 Step 3 — Lagrange Multiplier tests

Before choosing between a spatial lag or spatial error model, run the
**LM tests**. They tell you which specification fits better.

```{r}
#| label: lm-tests

lm_tests <- lm.RStests(ols, ncr_w, test = "all")
summary(lm_tests)
```

| Test | Significant? | Suggests |
|------------------------|------------------------|------------------------|
| **RSlag** | Yes | Spatial lag model (neighbours' rent affects own rent) |
| **RSerr** | Yes | Spatial error model (unobserved factors cluster spatially) |
| **adjRSlag** | Yes, adjRSerr No | Prefer spatial lag |
| **adjRSerr** | Yes, adjRSlag No | Prefer spatial error |

None of the tests are statistically significant — all p-values are well
above 0.05. This is consistent with the borderline residual Moran's I
result: with only 17 cities, the data do not provide strong enough
evidence to formally prefer either a spatial lag or spatial error
specification over OLS. We proceed with both spatial models as
robustness checks rather than as corrections to a confirmed misspecification.
In practice, this is a common situation in city-level analysis with small
n — the spatial models are still worth estimating and comparing via AIC,
but the results should be interpreted cautiously.

### 7.4 Step 4 — Spatial lag model

The **spatial lag model** adds the *average rent of neighbouring cities*
as an explanatory variable:

$$\ln(\text{rent}_i) = \alpha + \rho \sum_j w_{ij} \ln(\text{rent}_j) + \beta_1 \cdot \text{road\_dist}_i + \beta_2 \cdot \text{college}_i + \varepsilon_i$$

The new term $\rho \sum_j w_{ij} \ln(\text{rent}_j)$ is the **spatial
lag** — it captures how much of city *i*'s rent is explained by its
neighbours' rents.

```{r}
#| label: spatial-lag-model

slag <- lagsarlm(
  log(mean_rentval) ~ road_dist_km + pct_college_grad,
  data   = ncr_map,
  listw  = ncr_w
)
summary(slag)
```

The spatial lag model gives three key results. First, **rho (ρ) = −0.04**
with a p-value of 0.89 — the spatial lag term is essentially zero and
far from significant. Neighbours' rent does not predict own rent once
road distance and college graduation are controlled for. Second,
**college graduation remains significant** (p = 0.029): a one percentage
point increase in college graduates is associated with roughly 8.6%
higher rent. Third, **road distance is not significant** (p = 0.21),
though the negative sign is in the expected direction.

Crucially, the **AIC of the spatial lag model (6.73) is higher than OLS
(4.75)** — adding the spatial lag term made the model fit *worse*, not
better. This is consistent with the insignificant LM tests: there is no
meaningful spatial lag structure in this data at the city level.

### 7.5 Step 5 — Spatial error model

As an alternative, the **spatial error model** allows unobserved factors
(e.g. shared infrastructure, regional policy) to cluster spatially:

$$\ln(\text{rent}_i) = \alpha + \beta_1 \cdot \text{road\_dist}_i + \beta_2 \cdot \text{college}_i + u_i$$
$$u_i = \lambda \sum_j w_{ij} u_j + \varepsilon_i$$

```{r}
#| label: spatial-error-model

serr <- errorsarlm(
  log(mean_rentval) ~ road_dist_km + pct_college_grad,
  data  = ncr_map,
  listw = ncr_w
)
summary(serr)
```

The spatial error model tells a similar story. **Lambda (λ) = 0.29**
with a p-value of 0.31 — the spatial error term is positive (nearby
cities share unobserved characteristics) but not statistically
significant. **College graduation remains significant** (p = 0.018)
with a slightly larger coefficient than in the lag model: a one
percentage point increase in college graduates is associated with about
8.9% higher rent. Road distance again falls short of significance
(p = 0.20).

The **AIC of the spatial error model (6.21) is also higher than OLS
(4.75)**, confirming that adding spatial structure does not improve fit.


### 7.6 Compare models

```{r}
#| label: model-comparison

# AIC comparison — lower AIC = better fit
tibble(
  Model = c("OLS", "Spatial Lag", "Spatial Error"),
  AIC   = c(AIC(ols), AIC(slag), AIC(serr)),
  `Spatial ρ or λ` = c(NA,
                        slag$rho,
                        serr$lambda)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "Model fit comparison")
```

::: callout-note
**Choosing a model**\
Pick the spatial model with the lower AIC. If both spatial models have
substantially lower AIC than OLS, spatial structure matters. The
coefficient on `road_dist_km` should be interpreted as the *direct*
effect of road access on rent, holding neighbours' rents constant.
:::

OLS has the lowest AIC of the three models, confirming that spatial
dependence does not meaningfully improve fit for this dataset. The
spatial lag's **ρ = −0.04** is negligible — neighbours' rent adds
nothing once the covariates are controlled for. The spatial error's
**λ = 0.29** is larger and positive, hinting that unobserved factors
(shared infrastructure, policy environment) do cluster spatially across
NCR cities, but not strongly enough to justify the additional parameter
given only 17 observations.

------------------------------------------------------------------------

## 8 · Mapping Residuals

Even after adding spatial terms, it is good practice to map the
residuals. Persistent spatial clusters in the residuals suggest we are
still missing an important variable.

```{r}
#| label: map-residuals
#| fig-cap: "OLS vs spatial lag model residuals — NCR cities"

ncr_map <- ncr_map %>%
  mutate(
    resid_ols  = residuals(ols),
    resid_slag = residuals(slag)
  )

p_ols_resid <- ggplot(ncr_map) +
  geom_sf(aes(fill = resid_ols), color = "white", linewidth = 0.8) +
  scale_fill_distiller(palette = "RdBu", direction = 1,
                       name = "Residual") +
  labs(title = "OLS residuals") +
  theme_void(base_size = 10) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_slag_resid <- ggplot(ncr_map) +
  geom_sf(aes(fill = resid_slag), color = "white", linewidth = 0.8) +
  scale_fill_distiller(palette = "RdBu", direction = 1,
                       name = "Residual") +
  labs(title = "Spatial lag residuals") +
  theme_void(base_size = 10) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_ols_resid + p_slag_resid +
  plot_annotation(
    title   = "Residual Maps — Do spatial clusters remain?",
    subtitle = "Blue = over-predicted · Red = under-predicted",
    theme   = theme(plot.title = element_text(face = "bold", size = 12))
  )
```

::: callout-tip
**What to look for**\
If the OLS residual map shows obvious colour clusters (e.g. all southern
cities are red), the spatial lag model should reduce that clustering. If
the spatial lag map looks more random (no clear pattern), the model has
successfully absorbed the spatial spillovers.
:::

The two residual maps are almost indistinguishable — the spatial lag
model has not meaningfully redistributed the prediction errors. The
dominant pattern in both maps is a **north-south divide**: southern
cities (Muntinlupa, Parañaque) are consistently under-predicted (blue —
model thinks rent should be lower than it is), while several central
cities are over-predicted (red — model overshoots actual rent). This
persistent pattern suggests the missing variable is not a spatial
spillover between neighbours but something structural about the
**south corridor** — possibly the Alabang growth centre, proximity to
SLEX, or the concentration of gated subdivisions that inflate rent
independently of CBD distance or education levels.


## 9 · Key Takeaways

- **Spatial autocorrelation** means nearby cities are more similar than
  distant ones — a fundamental feature of urban data that OLS ignores
- **Moran's I** is the standard test: positive values confirm spatial
  clustering; a p-value below 0.05 means the pattern is not random.
  Rent and college graduation both show positive spatial autocorrelation
  in NCR, but Makati and San Juan are isolated high-value outliers
  rather than centres of a broader cluster
- **osmextract** downloads pre-built OSM extracts from Geofabrik into R —
  faster and more reliable than querying the live Overpass API directly
- **dodgr** converts road networks into graphs so we can compute actual
  travel distances — Caloocan has the largest gap between road and
  straight-line distance, reflecting poor north-south connectivity
- **Spatial lag and error models** are worth estimating even when LM
  tests are inconclusive — with n = 17, low statistical power means
  absence of significance is not absence of spatial structure
- **AIC, not just p-values, guides model selection**: here OLS
  outperformed both spatial models, validating the insignificant LM
  tests and suggesting spillovers operate below the city level
- **College graduation is the most robust predictor of rent** across all
  three model specifications — consistent with agglomeration and human
  capital externality theory
- A spatially patterned residual does not always call for a spatial
  model — sometimes it signals a missing variable, as with the
  south corridor pattern driven by Alabang and SLEX proximity

------------------------------------------------------------------------

## Session info

```{r}
#| label: session-info
#| code-fold: true

sessionInfo()
```
