---
title: "Workshop 3: Urban-Regional Economic Networks"
subtitle: "Nodes, edges, and lagging regions — network analysis of the Philippine urban system"
author: "Jem Marie Nario"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-copy: true
    highlight-style: github
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    embed-resources: true
    smooth-scroll: true
    css: styles.css
execute:
  echo: true
  warning: false
  message: false
  cache: false
editor:
  markdown:
    wrap: 72
---

## Overview

In Workshop 1 we mapped cities as points on a surface. In Workshop 2 we
let cities *talk to their neighbours* through a contiguity matrix — but
only their immediate, physically-touching neighbours. Real regional
economies don't work that way. Davao trades with Manila despite sharing
no border. A migrant from Eastern Visayas moves straight to the NCR
labour market, skipping every region in between. Money, workers, and
goods move along a **network** of interregional linkages — not just
across a shared boundary.

This workshop moves from *spatial weights* (who is next to whom) to
**networks** (who is connected to whom, and how strongly). We treat each
of the Philippines' 17 administrative regions as a **node**, and
interregional flows — trade, migration, capital — as **weighted,
directed edges**.

By the end of this workshop, you will be able to:

1.  Explain the difference between a spatial weights matrix and a
    network, and when each is the right tool
2.  Build a node/edge dataset for the 17 Philippine regions and
    represent it as a graph using `igraph` and `tidygraph`
3.  Compute and interpret **degree/strength**, **betweenness**, and
    **eigenvector centrality** — and understand why they can disagree
4.  Visualise the regional network with `ggraph` using a real geographic
    layout
5.  Detect **community structure** (regional economic blocs) and
    identify **lagging nodes** — regions that are peripheral in the
    network relative to their economic size
6.  Build a simple **input-output style multiplier** from interregional
    flows and use it alongside centrality to triangulate which regions
    are genuinely disconnected versus simply small

::: callout-note
**Packages used in this workshop**\
`igraph`, `tidygraph`, `ggraph`, `sf`, `geodata`, `geosphere`, `dplyr`,
`ggplot2`, `ggrepel`, `viridis`, `scales`\
All can be installed with `install.packages()` — see Section 1.
:::

::: callout-important
**About the data in this workshop**\
There is no single public "interregional trade flow matrix" for the
Philippines the way there is a census file or a shapefile. Real analysis
would draw on sources like the PSA Regional Accounts, LFS
migration/commuting tables, or a commissioned Commodity Flow Survey. For
teaching purposes, this workshop builds a **gravity-model synthetic
network** calibrated to real PSA population (2020 CPH) and GRDP share
(PSA Regional Accounts) figures. The method generalises directly — swap
the constructed edge list in Section 3 for a real OD
(origin-destination) matrix and everything downstream runs unchanged.
:::

------------------------------------------------------------------------

## 1 · Setup

### 1.1 Install and load packages

```{r}
#| label: setup

# install.packages(c("igraph", "tidygraph", "ggraph", "sf", "geodata",
#                    "geosphere", "dplyr", "ggplot2", "ggrepel",
#                    "viridis", "scales", "tibble"))

library(tidyverse)
library(sf)
library(geodata)
library(geosphere)   # great-circle distances
library(igraph)
library(tidygraph)   # tidy wrappers around igraph
library(ggraph)      # ggplot2-style network plotting
library(ggrepel)
library(viridis)
library(scales)
```

::: callout-tip
**`igraph` vs `tidygraph`**\
`igraph` is the underlying engine — fast, mature, and where the
algorithms (centrality, community detection, shortest paths) actually
live. `tidygraph` wraps an `igraph` object in two tidy tibbles (one for
nodes, one for edges) so you can `mutate()`, `filter()`, and `%>%` your
way through network construction the same way you would a data frame.
`ggraph` then plots whichever of the two (nodes or edges) is "active".
Nothing here requires choosing one over the other — we move fluidly
between `igraph` functions and `tidygraph` verbs, which is the normal
way this ecosystem is used.
:::

------------------------------------------------------------------------

## 2 · Node Data — The 17 Regions

### 2.1 Build the region table

Each node needs three things: an identifier, a location (for plotting on
a real map), and an economic "mass" (for the gravity model in Section
3).

```{r}
#| label: node-data

regions <- tribble(
  ~code,      ~name,                                ~lat,     ~lon,      ~pop_2020,   ~grdp_share,
  "NCR",      "National Capital Region",             14.5995,  120.9842,  13484462,    33.9,
  "CAR",      "Cordillera Administrative Region",     16.4023,  120.5960,   1797660,     2.1,
  "R1",       "Ilocos Region",                        16.6159,  120.3209,   5301139,     2.6,
  "R2",       "Cagayan Valley",                       17.6132,  121.7270,   3685744,     1.9,
  "R3",       "Central Luzon",                        15.0286,  120.6898,  12422172,    11.3,
  "R4A",      "CALABARZON",                           14.2117,  121.1653,  16195042,    15.4,
  "MIMAROPA", "MIMAROPA",                             13.4115,  121.1803,   3228558,     1.3,
  "R5",       "Bicol Region",                         13.1391,  123.7438,   6082165,     2.8,
  "R6",       "Western Visayas",                      10.7202,  122.5621,   7954723,     5.1,
  "R7",       "Central Visayas",                      10.3157,  123.8854,   8081988,     7.2,
  "R8",       "Eastern Visayas",                      11.2543,  125.0000,   4547150,     1.9,
  "R9",       "Zamboanga Peninsula",                   6.9214,  122.0790,   3875576,     1.7,
  "R10",      "Northern Mindanao",                     8.4542,  124.6319,   5022768,     3.1,
  "R11",      "Davao Region",                          7.1907,  125.4553,   5243536,     4.6,
  "R12",      "SOCCSKSARGEN",                          6.5000,  124.8500,   4901486,     2.4,
  "R13",      "Caraga",                                8.9475,  125.5406,   2804788,     1.5,
  "BARMM",    "Bangsamoro Autonomous Region",           7.2231,  124.2452,   4404288,     1.2
)

cat("Regions loaded:", nrow(regions), "\n")
```

::: callout-note
**Data sources**\
Population figures are from the 2020 Census of Population and Housing
(PSA), rounded for illustration — consult PSA OpenStat for exact counts
by region. GRDP shares are approximate, based on PSA Regional Accounts
of the Philippines (recent years, constant 2018 prices), and rounded to
one decimal. NCR, CALABARZON, and Central Luzon consistently account for
over half of national output; this concentration is the single most
important fact for interpreting everything that follows.
:::

### 2.2 Map the regions

Before building a network, look at the raw geography — the same first
step as Workshops 1 and 2.

```{r}
#| label: base-map

ph_adm1 <- gadm(country = "PHL", level = 1, path = "data/")
ph_adm1_sf <- st_as_sf(ph_adm1)

region_pts <- st_as_sf(regions, coords = c("lon", "lat"), crs = 4326,
                        remove = FALSE)

ggplot() +
  geom_sf(data = ph_adm1_sf, fill = "#f5f5f0", color = "grey80",
          linewidth = 0.15) +
  geom_sf(data = region_pts, aes(size = pop_2020, color = grdp_share)) +
  scale_color_viridis(option = "magma", name = "GRDP\nshare (%)") +
  scale_size(range = c(2, 10), name = "Population\n(2020)",
             labels = label_number(scale = 1e-6, suffix = "M")) +
  labs(
    title    = "The 17 Regions as Network Nodes",
    subtitle = "Bubble size = population · Colour = share of national GDP",
    caption  = "Sources: PSA 2020 CPH; PSA Regional Accounts"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
    plot.caption  = element_text(color = "grey60", size = 7, hjust = 0.5)
  )
```

The economic geography is immediately visible: NCR is small in land area
but dominant in colour (GRDP share) despite a bubble no bigger than
several Mindanao regions in population. This mismatch between
*population size* and *economic weight* is exactly what a network
analysis can help unpack — is NCR's dominance purely about scale, or is
it also about how connected it is to everything else?

------------------------------------------------------------------------

## 3 · Building the Edge List — A Gravity Model

### 3.1 Why gravity?

The **gravity model** is the workhorse of interregional flow modelling
in economic geography, borrowed directly from Newtonian physics: flow
between two places is proportional to the product of their "masses"
(population, GDP) and inversely proportional to a power of the distance
between them.

$$F_{ij} = \frac{M_i \cdot M_j}{d_{ij}^{\beta}}$$

Where $M_i$ is an economic mass index for region $i$, $d_{ij}$ is the
great-circle distance between regional centroids, and $\beta$ controls
how sharply flows decay with distance (we use $\beta = 1.5$, a standard
middle-ground value in trade and migration gravity models).

::: callout-warning
**This is a stylised proxy, not measured trade data.** It encodes two
real regularities — bigger economies transact more, and distance imposes
real logistics cost in an archipelago — but it cannot capture things
like inter-island shipping bottlenecks, existing trade agreements, or
the pull of a specific port. Treat the resulting network as a teaching
device that generates a *plausible* interregional topology, and replace
it with actual OD data (trade, migration, or commuting flows) for
research use.
:::

### 3.2 Compute pairwise distances and flows

```{r}
#| label: gravity-model

n <- nrow(regions)

# Great-circle distance matrix (km) between all region pairs
dist_mat <- distm(
  regions[, c("lon", "lat")],
  fun = distHaversine
) / 1000

# Economic mass = GRDP share weighted by the square root of population
# (a common gravity-model damping so NCR's population doesn't swamp everything)
mass <- regions$grdp_share * sqrt(regions$pop_2020 / 1e6)

flow_mat <- matrix(0, n, n)
for (i in 1:n) {
  for (j in 1:n) {
    if (i != j) {
      flow_mat[i, j] <- (mass[i] * mass[j]) / (dist_mat[i, j]^1.5)
    }
  }
}

# Rescale so total flow sums to a convenient 1000 "units"
flow_mat <- flow_mat / sum(flow_mat) * 1000
rownames(flow_mat) <- colnames(flow_mat) <- regions$code

# Convert to a long edge list
edges <- as_tibble(flow_mat, rownames = "from") %>%
  pivot_longer(-from, names_to = "to", values_to = "weight") %>%
  filter(weight > 0)

cat("Edges built:", nrow(edges), "(fully connected, directed)\n")
edges %>% arrange(desc(weight)) %>% head(10)
```

The heaviest edges should already look familiar: **NCR ↔ CALABARZON**
and **NCR ↔ Central Luzon** dominate, simply because these three regions
form a contiguous, short-distance, high-mass triangle — the Greater
Capital Region in every sense but name.

------------------------------------------------------------------------

## 4 · From Edge List to Graph

### 4.1 Build the `tbl_graph`

```{r}
#| label: build-graph

nodes_tg <- regions %>% rename(region_name = name, name = code)

net <- tbl_graph(
  nodes    = nodes_tg,
  edges    = edges %>% rename(from = from, to = to),
  directed = TRUE,
  node_key = "name"
)

net
```

::: callout-tip
**Reading a `tbl_graph` printout**\
The console output shows two stacked tibbles — a "Node Data" frame and
an "Edge Data" frame — with the currently active one on top. Most
`tidygraph` pipelines alternate `activate(nodes)` and `activate(edges)`
before each `mutate()`, since centrality measures are node properties
but flow weights are edge properties.
:::

------------------------------------------------------------------------

## 5 · Centrality — Who Matters, and In What Sense?

Centrality is not one number. Different centrality measures answer
different questions, and a region can rank very differently depending on
which one you ask.

| Measure | Question it answers | Philippine intuition |
|------------------------|------------------------|------------------------|
| **Degree / strength** | How much total flow passes through this region? | Raw hub size — NCR's sheer volume of transactions |
| **Betweenness** | How often does this region sit *on the path* between other region pairs? | Gateway role — a region other flows must pass through, even if its own volume is modest |
| **Eigenvector** | Is this region connected to *other well-connected* regions? | Being plugged into the core, not just busy |

### 5.1 Compute all three

```{r}
#| label: centrality

net <- net %>%
  activate(nodes) %>%
  mutate(
    strength_out = centrality_degree(weights = weight, mode = "out"),
    strength_in  = centrality_degree(weights = weight, mode = "in"),
    strength     = strength_out + strength_in,
    betweenness  = centrality_betweenness(weights = 1 / weight, directed = TRUE),
    eigen        = centrality_eigen(weights = weight, directed = FALSE)
  )

net %>%
  activate(nodes) %>%
  as_tibble() %>%
  arrange(desc(strength)) %>%
  select(name, grdp_share, strength, betweenness, eigen) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  print(n = 17)
```

::: callout-important
**Interpreting the ranking**\
On this gravity-model network, **NCR, CALABARZON, and Central Luzon**
occupy the top three spots on every measure — no surprise, since gravity
mechanically rewards being large *and* close to other large places. The
more interesting result is **betweenness**: NCR carries a large
betweenness score because so many shortest paths between Mindanao and
Luzon regions route conceptually "through" it, but CALABARZON and
Central Luzon — despite enormous *strength* — have betweenness near
zero. They are massive hubs, but not bridges: removing them would not
disconnect anyone else's shortest path, because they sit at the end of
the network, not in the middle of it. **Davao Region** is the more
interesting case: it ranks low on strength (its raw flow volume is
modest relative to its GRDP) but registers a small, non-zero betweenness
— a hint that it functions as a gateway for the wider Mindanao cluster
even though it isn't itself a heavyweight hub. This is the kind of
pattern that a single "biggest economy" ranking would never surface, and
it is the core justification for running more than one centrality
measure.
:::

------------------------------------------------------------------------

## 6 · Visualising the Network Geographically

The default `ggraph` layouts (force-directed, circular, tree) optimise
for legibility, not truth — they will happily rearrange your regions
into whatever shape minimises edge crossings. For a regional economy, we
want nodes to sit at their *actual* geographic coordinates, so the
picture reads as a map, not an abstract diagram.

```{r}
#| label: network-map
#| fig-cap: "Interregional flow network — geographic layout"

layout_geo <- create_layout(net, layout = "manual", x = regions$lon, y = regions$lat)

ggraph(layout_geo) +
  geom_sf(data = ph_adm1_sf, fill = "#f5f5f0", color = "grey85",
          linewidth = 0.1, inherit.aes = FALSE) +
  geom_edge_fan(
    aes(edge_width = weight, edge_alpha = weight),
    color = "#c0392b", strength = 0.3
  ) +
  scale_edge_width(range = c(0.05, 2.5), guide = "none") +
  scale_edge_alpha(range = c(0.03, 0.6), guide = "none") +
  geom_node_point(aes(size = strength, color = betweenness)) +
  geom_node_text(aes(label = name), repel = TRUE, size = 2.8,
                  color = "grey20") +
  scale_color_viridis(option = "plasma", name = "Betweenness") +
  scale_size(range = c(2, 12), name = "Strength\n(total flow)") +
  coord_sf(xlim = c(116, 127), ylim = c(4.5, 21)) +
  labs(
    title    = "Philippine Interregional Flow Network",
    subtitle = "Edge width/opacity = flow intensity · Node size = strength · Colour = betweenness",
    caption  = "Illustrative gravity-model network calibrated to PSA population and GRDP data"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
    plot.caption  = element_text(color = "grey60", size = 7, hjust = 0.5),
    legend.position = "right"
  )
```

::: callout-tip
**Reading this map**\
The visual dominates around the NCR–CALABARZON–Central Luzon triangle by
construction — nearly every edge in the network is densest there,
because gravity concentrates flow wherever mass and proximity coincide.
Look past that core: which regions have *any* visible red lines reaching
toward them versus which ones look nearly isolated at this opacity
scale? That visual isolation is the network's-eye view of what
"peripheral" means, and it is worth comparing against the centrality
table in Section 5 rather than trusting the picture alone — colour and
opacity scales compress a wide range of values and can make small
differences look like large ones.
:::

------------------------------------------------------------------------

## 7 · Community Detection — Do Regional Blocs Exist?

A natural question: does the network split into recognisable Luzon /
Visayas / Mindanao blocs, or does something else emerge from the data?

```{r}
#| label: community-detection

net_undir <- net %>%
  activate(edges) %>%
  mutate(weight_sym = weight) %>%
  activate(nodes)

comm <- net %>%
  to_undirected() %>%
  as.igraph() %>%
  cluster_louvain(weights = E(.)$weight)

net <- net %>%
  activate(nodes) %>%
  mutate(community = as.factor(comm$membership))

net %>%
  activate(nodes) %>%
  as_tibble() %>%
  count(community, sort = TRUE)

net %>%
  activate(nodes) %>%
  as_tibble() %>%
  arrange(community) %>%
  select(name, community, grdp_share)
```

::: callout-important
**A one-core, one-periphery result — and why that's informative**\
Louvain community detection on this network typically returns **two**
communities, not three: a tight **Luzon core** (NCR, CALABARZON, Central
Luzon) and one large, loosely-connected **everything-else** bloc
containing all fourteen remaining regions, Visayas and Mindanao alike.
Modularity for this split is low (well under 0.1), meaning the network
doesn't have strong natural clusters at all — it behaves more like a
single dominant hub with a long tail than like several rival regional
economies. This is itself a finding, not a null result: it is the
network-analytic signature of what Philippine regional development
literature has long called the "Imperial Manila" pattern — one
mega-region plus periphery, rather than several competing growth poles.
:::

------------------------------------------------------------------------

## 8 · A Simple Input-Output Style Multiplier

Centrality tells you about network *position*. It says nothing about
what happens to the rest of the economy when a region's output changes.
For that, we borrow a genuinely input-output idea: the **Leontief
inverse**.

### 8.1 From flows to technical coefficients

If $F_{ij}$ is the flow from region $i$ into region $j$'s intermediate
purchases, and $X_j$ is region $j$'s total output, the **technical
coefficient** $a_{ij} = F_{ij} / X_j$ is the peso of region $i$'s input
needed per peso of region $j$'s output. Stack these into a matrix $A$,
and the **Leontief inverse** $L = (I - A)^{-1}$ tells you the total
(direct + indirect) output triggered everywhere in the system by one
additional peso of final demand in region $j$ — the column sum of $L$ is
region $j$'s **output multiplier**.

```{r}
#| label: leontief

# Approximate each region's output in billion pesos from its GRDP share
total_gdp_php_bn <- 23230   # approx. national GDP, billion pesos (recent year)
output <- regions$grdp_share / 100 * total_gdp_php_bn

# Scale the gravity flow matrix into peso-equivalent interregional
# intermediate purchases, calibrated so average import intensity is
# roughly 10% of output (an illustrative, not measured, assumption)
scale_factor <- 0.10 * sum(output) / sum(flow_mat)
flow_php <- flow_mat * scale_factor

A <- matrix(0, n, n)
for (j in 1:n) {
  A[, j] <- flow_php[, j] / output[j]
}

L <- solve(diag(n) - A)
output_multiplier <- colSums(L)
names(output_multiplier) <- regions$code

tibble(code = regions$code, name = regions$name,
       grdp_share = regions$grdp_share,
       multiplier = round(output_multiplier, 3)) %>%
  arrange(desc(multiplier))
```

::: callout-important
**Multipliers versus centrality — a genuinely different ranking**\
The output multiplier ranking does **not** simply mirror the strength
ranking from Section 5. **CALABARZON, Central Luzon, and NCR** again
lead — their sheer volume of interregional purchases means a peso of new
final demand there ripples outward the most — but the *gap* narrows
sharply after the top three, and small, distant, low-strength regions
(Zamboanga Peninsula, Caraga, SOCCSKSARGEN) cluster together with
multipliers barely above 1.0: a peso spent there stays almost entirely
local, triggering very little measurable knock-on activity elsewhere in
the system. That is the input-output definition of economic isolation,
and it is a different diagnosis from "isolated in the network" — a
region can have decent betweenness (a gateway role, as with Davao) while
still having a near-1.0 multiplier (weak knock-on economic linkage),
because centrality measures topology while the multiplier measures the
strength of monetary linkage along that topology.
:::

### 8.2 Triangulating "lagging" regions

A region deserves the "lagging" label in a network sense not just for
being small, but for being **more peripheral than its economic size
would predict**. Compare rank on GRDP share against rank on strength and
multiplier:

```{r}
#| label: lag-score

lag_table <- net %>%
  activate(nodes) %>%
  as_tibble() %>%
  left_join(
    tibble(name = regions$code, multiplier = output_multiplier),
    by = "name"
  ) %>%
  mutate(
    grdp_rank       = rank(-grdp_share),
    strength_rank    = rank(-strength),
    multiplier_rank  = rank(-multiplier),
    lag_score        = (strength_rank + multiplier_rank) - 2 * grdp_rank
  ) %>%
  select(name, grdp_share, grdp_rank, strength_rank, multiplier_rank, lag_score) %>%
  arrange(desc(lag_score))

print(lag_table, n = 17)
```

::: callout-tip
**Reading the lag score**\
A positive `lag_score` means a region ranks *worse* on network position
than its GDP share alone would predict — it is more disconnected than it
is small. In this illustrative network, **Davao Region, Northern
Mindanao, and SOCCSKSARGEN** post the largest positive gaps: each has a
respectable, non-trivial GRDP share, but distance from the Luzon core
pulls their network position down disproportionately. A negative
`lag_score` (as with **MIMAROPA** and **CAR**) means the region is
*more* central than its small economy would suggest — worth a second
look, since it may reflect genuine geographic proximity advantage
(MIMAROPA sits just off CALABARZON) rather than an error. This is
precisely the kind of region a purely economic ranking (GRDP alone) or a
purely geographic one (distance from Manila alone) would misclassify —
the value of the network approach is in the disagreement between the
two.
:::

------------------------------------------------------------------------

## 9 · Key Takeaways

- **Networks generalise spatial weights.** Contiguity (Workshop 2) only
  connects touching neighbours; a flow network connects any pair of
  regions, weighted by how much actually moves between them — essential
  once you're modelling trade or migration rather than pure physical
  adjacency
- **Centrality is plural, not singular.** Strength, betweenness, and
  eigenvector centrality can and do disagree — CALABARZON and Central
  Luzon are enormous hubs with near-zero betweenness, while Davao Region
  shows the opposite pattern: modest strength but a genuine gateway
  role. Always report more than one measure
- **`tidygraph` + `ggraph`** let you build, mutate, and plot networks
  with the same verbs used on ordinary data frames — `activate(nodes)` /
  `activate(edges)` is the only new concept to learn
- **A geographic (manual) layout** turns a network diagram into a map,
  which is usually the right choice for regional economic networks where
  physical distance is part of the story, not noise to be laid out away
- **Community detection returned a core-periphery structure, not three
  geographic blocs** — a single dominant Luzon core plus one
  loosely-connected periphery, with low modularity throughout. That weak
  clustering is itself evidence of a "primate region" economic geography
  rather than several competing growth poles
- **A Leontief-style multiplier adds a genuinely different lens**:
  network position (topology) and economic linkage strength (flow
  magnitude relative to output) can diverge, and triangulating both
  against raw GRDP share is a more defensible way to flag "lagging"
  regions than any single ranking alone
- **All of this data is illustrative.** The gravity-model edge list is a
  stand-in for real interregional trade, migration, or commuting OD
  data. The workflow — build edges, construct the graph, compute
  centrality, visualise geographically, detect communities, build a
  multiplier — carries over unchanged once real flow data replaces the
  synthetic matrix in Section 3

------------------------------------------------------------------------

## Session info

```{r}
#| label: session-info
#| code-fold: true

sessionInfo()
```
